---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 3,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rapr

<!-- badges: start -->
[![CRAN
version](https://www.r-pkg.org/badges/version/rapr)](https://CRAN.R-project.org/package=rapr)
[![CRAN
status](https://badges.cranchecks.info/worst/rapr.svg)](https://cran.r-project.org/web/checks/check_results_rapr.html)
[![Development
Version](https://brownag.r-universe.dev/badges/rapr)](https://brownag.r-universe.dev/)
[![html-docs](https://img.shields.io/badge/docs-HTML-informational)](https://humus.rocks/rapr/)
[![R-CMD-check](https://github.com/brownag/rapr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/brownag/rapr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/brownag/rapr/branch/main/graph/badge.svg)](https://app.codecov.io/gh/brownag/rapr?branch=main)
<!-- badges: end -->

The goal of {rapr} is to provide a simple **R** interface to 'Rangeland Analysis
Platform' (RAP) products. See https://rangelands.app/products and
`citation("rapr")` for more information.

## Installation

You can install {rapr} from CRAN:

``` r
install.packages("rapr")
```

You can install the development version of {rapr} from
[GitHub](https://github.com/brownag/rapr) with:

``` r
# install.packages("remotes")
remotes::install_github("brownag/rapr")
```

## Sources, Products, and Band Information

Two sets of gridded RAP products are available (see `get_rap()` `source` argument). You can also access four time series API endpoints providing tabular data (see `get_rap_table()` `product` argument).

 - `"rap-30m"` is Landsat-derived and has approximately 30 meter resolution in WGS84 decimal degrees (`"EPSG:4326"`). This is the data source that has been used in the 'rapr' package since 2022.

 - `"rap-10m"` is Sentinel 2-derived and has 10 meter resolution in the local WGS84 UTM zone (`"EPSG:326XX"`, where XX is the two digit UTM zone number). It is a newer (2025) data source.

For `"rap-30m"` you can query several Landsat derived annual biomass, cover, and
Net Primary Productivity products from 1986 to present:

  - `product = "vegetation-biomass"` returns [two layers](http://rangeland.ntsg.umt.edu/data/rap/rap-vegetation-biomass/v3/README) per year:

    - 2 Bands:`"annual forb and grass"`, `"perennial forb and grass"` (**lbs / acre**)

  - `product = "vegetation-cover"` returns [six layers](http://rangeland.ntsg.umt.edu/data/rap/rap-vegetation-cover/v3/README) per year:

    - 6 Bands: `"annual forb and grass"`, `"bare ground"`, `"litter"`, `"perennial forb and grass"`, `"shrub"`, `"tree"` (**% cover**)

  - `product = "vegetation-npp"` returns [four layers](http://rangeland.ntsg.umt.edu/data/rap/rap-vegetation-npp/v3/README) per year:

    - 4 Bands: `"annual forb and grass"`, `"perennial forb and grass"`, `"shrub"`, `"tree"` (NPP; kg*C/m^2)

For `"rap-10m"` you can query several [Sentinel 2 derived cover
products](http://rangeland.ntsg.umt.edu/data/rangeland-s2/README) at 10 meter
resolution from 2018 to present:

   - `product = "pft"` returns fractional cover estimates of plant functional types:

     - 6 Bands: `"annual forb and grass"`, `"bare ground"`, `"litter"`, `"perennial forb and grass"`, `"shrub"`, `"tree"`  (**% cover**)

   - `product = "gap"` returns canopy gap estimates for four canopy gap size classes:

     - 4 Bands: `"Gaps 25-50 cm"`, `"Gaps 51-100 cm"`, `"Gaps 100-200 cm"`, `"Gaps >200 cm"`  (**% cover**)

   - `product = "arte"` returns cover estimates of Artemisia species, including A. arbuscula, A. cana, A. nova, A. tridentata, and A. tripartita.

     - 1 Band: `"Artemisia spp."` (**% cover**)

   - `product = "iag"` returns fractional cover estimates of Bromus tectorum, B. arvensis, B. rubens, B. hordeaceus, Eremopyrum triticeum, Schismus spp., Taeniatherum caput-medusae, and Ventenata dubia.

     - 1 Band: `"invasive annual grass"` (**% cover**)

   - `product = "pj"` returns fractional cover estimates of Juniperus monosperma, J. occidentalis, J. osteosperma, J. scopulorum, Pinus edulis, and P. monophylla.

     - 1 Band: `"pinyon-juniper"` (**% cover**)
     
The four time series API endpoints accessed through `get_rap_table()` return mean values for each period and area of interest for a number of vegetation metrics:

   - `product = "cover"` returns yearly fractional cover, including:
       - `"AFG"` (Annual Forb and Grass cover)
       - `"PFG"` (Perennial Forb and Grass cover)
       - `"SHR"` (Shrub cover)
       - `"TRE"` (Tree cover) 
       - `"LTR"` (Litter cover)  
       - `"BGR"` (Bare Ground cover)
  
   - `product = "coverMeteorology"` returns the same data provided by `"cover"` above, plus:
       - `"annualTemp"` (Annual average temperature in degrees Fahrenheit) 
       - `"annualPrecip"` (Annual total precipitation in inches)
       
   - `product = "production"` returns annual production, including: 
       - `"AFG"` (Annual Forb and Grass production)
       - `"PFG"` (Perennial Forb and Grass production) 
       - `"HER"` (Herbaceous production)
  
   - `product = "production16day"` returns 16-day production, including:
       - `"date"` (production estimate date)
       - `"doy"` (production estimate Julian day of year)
       - `"AFG"` (Annual Forb and Grass production)
       - `"PFG"` (Perennial Forb and Grass production) 
       - `"HER"` (Herbaceous production)

Each of these time series is returned as a _data.frame_ with requested time-series data by year or 16-day production period. In addition to the columns described above, all tables include columns for `"year"` (production or cover estimate year) and `"feature"` (feature ID, row number from `aoi`). Units are **% cover** for fractional cover and **lbs / acre** for production.

## Temporary Files

  Large requests may generate intermediate objects that will be stored as
  temporary files. See `terra::tmpFiles()` to view the file paths. These
  files will be removed when an **R** session ends.

## Alternate Specification of Area of Interest

  In lieu of a spatial object from \{terra\}, \{raster\}, \{sf\} or \{sp\}
  packages you may specify a bounding box using a numeric vector containing
  the top-left and bottom-right coordinates (`xmin`, `ymax`, `xmax`, `ymin`)
  in WGS84 longitude/latitude decimal degrees. This corresponds to the
  conventional order used in the `gdal_translate` `-projwin` option. e.g.
  `get_rap(x = c(-120, 37, -119.99, 36.99), ...)`.

```
(1: xmin, 2: ymax)--------------------------|
        |                                   |
        |         TARGET EXTENT             |
        |  x = c(xmin, ymax, xmax, ymin)    |
        |                                   |
        |---------------------------(3: xmax, 4: ymin)
```
## Native Resolution and Projection Systems

  Native cell resolution of `"rap-30m"` is approximately 30m x 30m in WGS84
  geographic coordinate system (longitude, latitude). Native cell resolution of
  `"rap-10m"` is 10m x 10m in the local (projected) WGS84 Universal Transverse
  Mercator (UTM) system.

  For `"rap-10m"` requests spanning _multiple_ UTM zones, either pass a
  _SpatRaster_ object as `x` or specify `template` argument. In lieu of a
  user-specified grid system for multi-zone requests, a default CONUS Albers
  Equal Area projection (`"EPSG:5070"`) with 10 m resolution will be used. See
  `rap_projection()` for options and details.
  
## Example

This example shows how to use a {[terra](https://github.com/rspatial/terra)}
SpatVector containing a rectangular polygon with `rapr::get_rap()` to obtain RAP
grids and a mean annual vegetation cover table for the corresponding extent.
{[terra](https://cran.r-project.org/package=terra)},
{[raster](https://cran.r-project.org/package=raster)},
{[sf](https://cran.r-project.org/package=sf)} and
{[sp](https://cran.r-project.org/package=sp)} objects are all supported input
types. Any spatial object used should have its Coordinate Reference System
defined, as the input coordinates will be projected to the source data grid
system.

```{r example}
library(terra)
library(rapr)

res <- get_rap(
  vect("POLYGON ((-120 36.99,-119.99 37,-120 37,-120 36.99))",
       crs = "EPSG:4326"), 
  version = "v3",
  product = "vegetation-cover",
  years = c(1986, 1996, 2006, 2016),
  verbose = TRUE
)

res

plot(res)

# Mean fractional vegetation cover estimates
res_tab <- get_rap_table(
  vect("POLYGON ((-120 36.99,-119.99 37,-120 37,-120 36.99))",
       crs = "EPSG:4326"), 
  product = "cover",
  years = c(1989, 1999, 2009, 2019)
)

res_tab
```

In lieu of a spatial object from {terra}, {raster}, {sf} or {sp} packages you
may specify a bounding box using a numeric vector containing `xmin`, `ymax`,
`xmax`, `ymin` in WGS84 longitude/latitude decimal degrees (corresponding to
order used in `gdal_translate` `-projwin` option). e.g. `get_rap(x = c(-120, 37,
-119.99, 36.99), ...)`.

```
(1: xmin, 2: ymax)--------------------------|
        |                                   |
        |         TARGET EXTENT             |
        |  x = c(xmin, ymax, xmax, ymin)    |
        |                                   |
        |---------------------------(3: xmax, 4: ymin)
```

## Citation

```{r, echo=FALSE}
citation("rapr")
```
